---
title: "Google Data Analytics Capstone Project - Cyclistic Analysis"
author: "Kevin Collado"
date: "2022-08-11"
output:
  html_document: default
  pdf_document: default
---

# [Cyclistic Full Year Analysis]{.underline}

## setup: load packages

```{r setup, include=FALSE}
library(tidyverse)    #for data cleaning
library(dplyr)        #for data manipulation
library(lubridate)    #for date parsing and manipulation
library(ggplot2)      #data viewing
library(readxl)       #to load excel files
setwd(getwd())
```

# STEP 1: COLLECT DATA

```{r}
setwd("~/Documents/Google Data Analytics Course/Capstone Data XLS")
bikeshare_2021_08 <- read_xls("2021-08-divvy-tripdata.xls")
bikeshare_2021_09 <- read_xls("2021-09-divvy-tripdata.xls")
bikeshare_2021_10 <- read_xls("2021-10-divvy-tripdata.xls")
bikeshare_2021_11 <- read_xls("2021-11-divvy-tripdata.xls")
bikeshare_2021_12 <- read_xls("2021-12-divvy-tripdata.xls")
bikeshare_2022_01 <- read_xls("2022-01-divvy-tripdata.xls")
bikeshare_2022_02 <- read_xls("2022-02-divvy-tripdata.xls")
bikeshare_2022_03 <- read_xls("2022-03-divvy-tripdata.xls")
bikeshare_2022_04 <- read_xls("2022-04-divvy-tripdata.xls")
bikeshare_2022_05 <- read_xls("2022-05-divvy-tripdata.xls")
bikeshare_2022_06 <- read_xls("2022-06-divvy-tripdata.xls")
bikeshare_2022_07 <- read_xls("2022-07-divvy-tripdata.xls")
```

# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

## compare column names and check for consistency

```{r}
colnames(bikeshare_2021_08)
colnames(bikeshare_2021_09)
colnames(bikeshare_2021_10)
colnames(bikeshare_2021_11)
colnames(bikeshare_2021_12)
colnames(bikeshare_2022_02)
colnames(bikeshare_2022_03)
colnames(bikeshare_2022_04)
colnames(bikeshare_2022_05)
colnames(bikeshare_2022_06)
colnames(bikeshare_2022_07)
```

## rename columns for consistency

```{r}
bikeshare_2021_08 <- rename(bikeshare_2021_08, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2021_09 <- rename(bikeshare_2021_09, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2021_10 <- rename(bikeshare_2021_10, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2021_11 <- rename(bikeshare_2021_11, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2021_12 <- rename(bikeshare_2021_12, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_01 <- rename(bikeshare_2022_01, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_02 <- rename(bikeshare_2022_02, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_03 <- rename(bikeshare_2022_03, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_04 <- rename(bikeshare_2022_04, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_05 <- rename(bikeshare_2022_05, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_06 <- rename(bikeshare_2022_06, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)

bikeshare_2022_07 <- rename(bikeshare_2022_07, trip_id = ride_id, bikeid = rideable_type, start_time = started_at, end_time = ended_at, from_station_name = start_station_name, from_station_id = start_station_id, to_station_name = end_station_name, to_station_id = end_station_id, usertype = member_casual)
```

## inspect dataframes for incongruencies

```{r}
str(bikeshare_2021_08)
str(bikeshare_2021_09)
str(bikeshare_2021_10)
str(bikeshare_2021_11)
str(bikeshare_2021_12)
str(bikeshare_2022_01)
str(bikeshare_2022_02)
str(bikeshare_2022_03)
str(bikeshare_2022_04)
str(bikeshare_2022_05)
str(bikeshare_2022_06)
str(bikeshare_2022_07)
```

## convert trip_id and bikeid to chr type to ensure stacking

```{r}
bikeshare_2021_08 <- mutate(bikeshare_2021_08, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2021_09 <- mutate(bikeshare_2021_09, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2021_10 <- mutate(bikeshare_2021_10, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2021_11 <- mutate(bikeshare_2021_11, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2021_12 <- mutate(bikeshare_2021_12, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_01 <- mutate(bikeshare_2022_01, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_02 <- mutate(bikeshare_2022_02, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_03 <- mutate(bikeshare_2022_03, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_04 <- mutate(bikeshare_2022_04, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_05 <- mutate(bikeshare_2022_05, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_06 <- mutate(bikeshare_2022_06, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
bikeshare_2022_07 <- mutate(bikeshare_2022_07, trip_id = as.character(trip_id), bikeid = as.character(bikeid))
```

## stack all data frames into one dataframe

```{r}
all_trips <- bind_rows(bikeshare_2021_08, bikeshare_2021_09, bikeshare_2021_10, bikeshare_2021_11, bikeshare_2021_12, bikeshare_2022_01, bikeshare_2022_02, bikeshare_2022_03, bikeshare_2022_04, bikeshare_2022_05, bikeshare_2022_06, bikeshare_2022_07)
```

## remove columns that will not be used

```{r}
all_trips <- all_trips %>% select(-c(start_lat,start_lng,end_lat,end_lng))
```

# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

## inspect new tables

```{r}
colnames(all_trips)   #check list of column names
nrow(all_trips)       #check number of rows
dim(all_trips)        #check dimensions
head(all_trips)       #check first 6 rows
tail(all_trips)       #check last 6 rows
str(all_trips)        #check list of columns and data types
summary(all_trips)    #retrieve statistical summary
```

## date manipulation

```{r}
all_trips$date <- as.Date(all_trips$start_time)
all_trips$month <- format(as.Date(all_trips$date),"%m")
all_trips$day <- format(as.Date(all_trips$date),"%d")
all_trips$year <- format(as.Date(all_trips$date),"%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date),"%A")
```

## add ride_length data with start and end times

```{r}
all_trips$ride_length <- difftime(all_trips$end_time,all_trips$start_time)
```

## re-inspect structure of columns

```{r}
str(all_trips)
```

## ensure ride_length is numeric to allow for calculations

```{r}
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

## create a version with no negative ride_length values

```{r}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR"|all_trips$ride_length<0),]
```

# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

```{r}
summary(all_trips_v2$ride_length)
```

## compare members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = min)
# then compare average ride times by each day for members vs casual members
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
# reorder days of week
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
# check average ride times by each day again
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
```

# analyze ridership data by type and weekday

```{r}
all_trips_v2 %>% mutate(weekday=wday(start_time, label=TRUE)) %>% group_by(usertype,weekday) %>% summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% arrange(usertype,weekday)
```

## visualize the number of rides by rider type

```{r}
all_trips_v3 <- all_trips_v2[!is.na(all_trips_v2$ride_length),]
# created a version without NA ride length values
all_trips_v3 %>% mutate(weekday=wday(start_time,label=TRUE)) %>% group_by(usertype,weekday) %>% summarise(number_of_rides=n(),number_of_rides=mean(ride_length)) %>% ggplot(aes(x=weekday,y=number_of_rides,fill=usertype)) + geom_col(position="dodge")
```

## then visualize for average duration

```{r}
all_trips_v3 %>% mutate(weekday=wday(start_time,label=TRUE)) %>% group_by(usertype,weekday) %>% summarise(number_of_rides=n(),average_duration=mean(ride_length)) %>% ggplot(aes(x=weekday,y=average_duration,fill=usertype)) + geom_col(position="dodge")
```

#STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS \## create csv files to be visualized in Tableau

```{r}
# save main data
write.csv(all_trips_v2,file="/Users/kevincollado/Documents/Google Data Analytics Course/Capstone Data CSV/all_trips_v2.csv")
# then save a csv file containing only average ride length, usertype, and day of week
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts,file="/Users/kevincollado/Documents/Google Data Analytics Course/Capstone Data CSV/avg_ride_length.csv")
```

# Finished!
